Stemmer Tool Manual

### 2018-03-14 tex/merf/OldSubmission/gui.tex
old submission files

\framework~provides a user friendly interface to specify the 
atomic terms, the MBFs, the MREs, 
the tag types, and the legends. 
The GUI also allows the user to modify and correct the 
tag set $R$. 
The GUI allows the user also to compute accuracy results 
that compare different tag sets and that can serve well as 
inter annotation agreement results
when the tag sets come from two human annotators, 
or as evaluation results when comparing with reference tag sets.

\framework saves the tags and the tag types in a user friendly format using
the JavaScript Object Notation (JSON) format~\cite{nolan2014javascript}.
The tag file keeps the paths of the separate text and tag type files. 
It also contains a list of tags with their tag type identifiers,
position in text, length, and word index.
The tag type file contains MBF and MRE tag types. 
%Each object has an array value containing the user-defined tag types. 
Each tag type contains the name, description, expression, and the visualization legends.

%Figure~\ref{fig:mbftagger} shows the \framework GUI. 
%%\framework's GUI allows the user to create, manipulate, and analyze a project. 
%The {\em File} menu is used to create, close, or open an existing project. 
%The {\em Tags} menu enables the user to manually define and edit tag types. 
%The user can define and edit the MBF and MRE based tag types and 
%start the corresponding automatic morphology-based simulators from the {\em Tagtypes} menu. 
%The user performs comparison and analysis of two tag sets using the {\em Analyse} menu. 
%{\em View} menu enables the user to switch visualization modes of the tags.

\subsection{Tag type Boolean formula editor}
The user writes MBF tag types with the tag type editor introduced in~\cite{JaZaMatar}. %shown in Figure~\ref{f:bfe}.
First the user specifies atomic terms by selecting a feature from ${\cal F}$. 
%The pattern filters the feature values.
The user can also choose whether to require an exact match using the
\cci{isA} predicate, or a substring match using the
\cci{contains} predicate option.

The user can add and remove feature values to the atomic terms 
using push buttons. 
A check box in the ``Feature'' column allows negating the term, and the
``Relation'' column switches the predicate between 
\cci{isA} and \cci{contains}. 
The list of feature and value pairs is interpreted as a disjunction to form
the MBF. 
A right pane shows a description of the tag type and a set of legend 
descriptors. 
When the stem or gloss features are selected, the user has the option to 
use the $Syn^k$ feature. 
%The user selects the check box in the lower left of the editor and sets the order in the spin box.

In the direction extraction task example, the user specifies four MBF-based 
tag types with labels 
$N$, $P$, $R$, and $U$ with  ``name of person'', ``name of place'', 
``relative position'', and ``numerical term'' descriptions, respectively. 
For each MBF, the user selects the morphological features, 
specifies the constant value $CF$, and adds it to the Boolean formula editor. 
%The user also assigns legend descriptors to each defined tag type. 
%
%Figure~\ref{f:bfe} shows the $U$ MBF which requires the gloss tag of 
%the stem to be an exact match of either {\tt first}, {\tt second}, \dots, or {\tt tenth}. 
%The legend descriptors set foreground and background colors to {\tt orangered}, 
%and {\tt gray}, respectively, and the font size to {\tt 12}.
%, normal font weight, and unitalicized.

\begin{figure}[tb!]
  \centering
  {\setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0.5pt}%
  \fbox{\includegraphics[width=\textwidth]{figures/mreedit}}}
  \caption{\framework tag type regular expression editor.}
  \label{f:sfe}
\end{figure}

@@

%After defining the MBFs of the direction extraction task, 
%the user calls the MBF simulator from the {\tt Tagtypes} menu. 
%The snapshot in Figure~\ref{fig:mbftagger} shows 
The MBF match visualizer shows color sensitive text view, the tag list view, and the tag description view. 
%The color-sensitive text view shows the text with visualization of the MBF tag type matches. 
%The tag list view shows all tags.
%that are automatically or manually applied to the text. 
The tag description view presents the details of the selected tag along with 
the relevant tag type information.
%
The user can edit the tags using a context sensitive menus. 
\framework GUI also allows manual tag types and corresponding tags 
that are not based on morphological features.
This enables building reference corpora without 
help from the morphological analyzer.

\vspace{-1em}
\subsection{Tag type regular expression editor}

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=\textwidth]{figures/treegraph}
  \caption{MRE annotated Text, MRE match tree, and entity-relation graph}
  \label{fig:treegraph}
\end{figure}

\vspace{-1em}
\subsection{MRE match visualization}

While specifying an MRE the user can interact with the visualization and editor views
to make sure the MRE expresses the intent. 
The color-sensitive text view in Figure~\ref{fig:treegraph} shows 
the highlighted tag matches after the user called the MRE simulator using 
the {\tt Tagtypes} menu.

The match tree view shows the selected match in a graph view.
Figure~\ref{fig:treegraph} shows the MRE match tree of the direction task 
\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building). 
%Leaf nodes show the match text, internal nodes show the operations, and edge labels shows the MBF tag type name.

%\begin{figure}[tb]
%  \centering
%  \includegraphics[width=0.5\textwidth]{figures/merfsreditor.png}
%\caption{\framework~relation editor}
%  \label{fig:relationeditor}
%\end{figure}

\subsection{User defined relation editor}

After the user is satisfied with the MRE matches, 
the user moves to define relations and code actions. 
The relation editor allows the user to define relations 
by specifying $\langle e_1,e_2,r\rangle$ tuples, 
where $e_1$ and $e_2$ denote source and destination entities, and $r$ denotes 
the label.
The editor shows the MRE tree and allows the user to select the sub 
expressions and select features 
of the matches of the sub expressions to define the three components of the relation.

A snapshot of the GUI in 
Figure~\ref{fig:treegraph} shows in an interactive graph view
the entity-relation graph of the match of the user defined relation 
extracted from the match tree of the MRE.

>>>>

\begin{figure*}[tb!]
  \centering
  {\setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0.5pt}%
  \fbox{\includegraphics[width=0.8\textwidth]{figures/mbfedit}}}
  \caption{\framework tag type Boolean formula editor.}
  \label{f:bfe}
\end{figure*}

\begin{figure*}[tb!]
  \centering
  \includegraphics[width=0.85\textwidth]{figures/mbfmatchvis}
  \caption{\framework main window with MBF match annotated text, tag type properties, and manual tag editing menus.}
  \label{fig:mbftagger}
\end{figure*}

%In this Section, we present the user friendly interface of \framework. 
%We show how to build the direction extraction task shown in 
%Section~\ref{sec:motivation}. 
%The task requires the user to use the \framework interactively to 
%define Boolean formulae, define regular expressions, 
%declare the semantic relations.

\framework~provides a user friendly interface to specify the 
atomic terms, the \framework~Boolean formulae, the~\framework regular expressions, 
the tag types, and the legends. 
The \framework~GUI also allows the user to modify and correct the 
tag set $R$. 
The \framework~GUI allows the user also to compute accuracy results 
that compare different tag sets and that can sever well as 
inter annotation agreement results
when the tag sets come from two human annotators, 
or as evaluation results when comparing with reference tag sets.

\framework saves the tags and the tag types in a user friendly format using
the JavaScript Object Notation (JSON) data-interchange 
format~\cite{nolan2014javascript}.
The tag file keeps the paths of the separate text and tag type files. 
It also contains a list of tags with their tag type identifiers,
position in text, length, and word index.
The tag type file contains MBF and MRE tag types. 
%Each object has an array value containing the user-defined tag types. 
Each tag type contains the name, description, expression, and the visualization legends.

Figure~\ref{fig:mbftagger} shows the \framework GUI. 
%\framework's GUI allows the user to create, manipulate, and analyze a project. 
The {\em File} menu is used to create, close, or open an existing project. 
The {\em Tags} menu enables the user to manually define and edit tag types. 
The user can define and edit the MBF and MRE based tag types and 
start the correspoding automatic morphology-based simulators from the {\em Tagtypes} menu. 
The user performs comparison and analysis of two tag sets using the {\em Analyse} menu. 
{\em View} menu enables the user to switch visualization modes of the tags.

\subsection{Tag type Boolean formula editor}
The user writes MBF tag types with the tag type editor shown in Figure~\ref{f:bfe}.
First the user specifies atomic terms by selecting a feature from ${\cal F}$. 
The pattern filters the feature values.
The user can also chose whether to require an exact match using the
\cci{isA} predicate from the combobox, or a substring match using the
\cci{contains} predicate option.

The user can add and remove selected feature values to the atomic terms 
using the push buttons. 
The check box in the ``Feature'' column allows negating the term, and the
``Relation'' column menu switches the predicate between 
\cci{isA} and \cci{contains}. 
The list of feature and value pairs is interpreted as a disjunction to form
the MBF of the tag type. 
The right pane shows a description of the tag type and a set of legend 
descriptors. 
When the stem or gloss features are selected, the user has the option to 
use the $Syn^k$ feature. 
%The user selects the check box in the lower left of the editor and sets the order in the spin box.

In the direction extraction task example, the user specifies four MBF-based 
tag types with labels 
$N$, $P$, $R$, and $U$ with  ``name of person'', ``name of place'', 
``relative position'', and ``numerical term'' descriptions, respectively. 
For each MBF, the user selects the morphological features, 
specifies the constant value $CF$, and adds it to the Boolean formula editor. 
%The user also assigns legend descriptors to each defined tag type. 
%
Figure~\ref{f:bfe} shows the $U$ MBF which requires the gloss tag of 
the stem to be an exact match of either {\tt first}, {\tt second}, \dots, or {\tt tenth}. 
The legend descriptors set foreground and background colors to {\tt orangered}, 
and {\tt gray}, respectively, and the font size to {\tt 12}.
%, normal font weight, and unitalicized.

@@

After defining the MBFs of the direction extraction task, 
the user calls the MBF simulator from the {\tt Tagtypes} menu. 
The snapshot in Figure~\ref{fig:mbftagger} shows 
the color sensitive text view, the tag list view, and the tag description view. 
%The color-sensitive text view shows the text with visualization of the MBF tag type matches. 
The tag list view shows all tags.
%that are automatically or manually applied to the text. 
The tag description view presents the details of the selected tag along with 
the relevant tag type information.
%
The context sensitive menus allow the user to tag 
a selected word or chunk differently or to entirely remove the tag. 
\framework GUI also allows manual tag types and corresponding tags 
that are not based on morphological features.
This enables building reference corpora without 
help from the morphological analyzer.

\begin{figure*}[t]
  \centering
  {\setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0.5pt}%
  \fbox{\includegraphics[width=0.8\textwidth]{figures/mreedit}}}
  \caption{\framework tag type regular expression editor.}
  \label{f:sfe}
\end{figure*}

\subsection{Tag type regular expression editor}

@@

\begin{figure*}[tb!]
  \centering
  \includegraphics[width=\textwidth]{figures/treegraph}
%  \begin{tabular}{ccc}
%  \textbf{\relsize{-2} (a) Arabic text{\color{white}{....................................}}} &
%  \textbf{\relsize{-2} (b) Match Tree{\color{white}{....................................}}} &
%  \textbf{\relsize{-2} (c) Entity-relation graph} \\
%  \end{tabular}
  \caption{MRE annotated Text, MRE match tree, and entity-relation graph}
  \label{fig:treegraph}
\end{figure*}

\subsection{MRE match visualization}

While specifying an MRE the user can interact with the visualization and editor views
to make sure the MRE expresses the intent. 
%The user calls the MRE simulator from the {\tt Tagtypes} menu to compute the matches of the MRE expressions. 
The color-sensitive text view in Figure~\ref{fig:treegraph} shows 
the highlighted tag matches after the user called the MRE simulator using 
the {\tt Tagtypes} menu.

%The tag list view shows all the tag matches that are automatically or manually applied to the text. 
The match tree view shows the selected match in a graph view.
%tree view using 
%graphviz library. 
%This library provides a variety of software for drawing attributed graphs and computes the layout using a set of common graph layout algorithms such as dot. 
%We use the generated coordinates of the nodes in the graphviz graph and Qt's graphics items to draw the match tree in \framework's main window.
%
Figure~\ref{fig:treegraph} shows the match tree of the direction task 
regular expression match 
\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building). 
%Leaf nodes show the match text, internal nodes show the operations, and edge labels shows the MBF tag type name.

\begin{figure}[tb]
  \centering
  \includegraphics[width=0.5\textwidth]{figures/merfsreditor.png}
\caption{\framework~relation editor}
  \label{fig:relationeditor}
  \vspace{-1em}
\end{figure}

\subsection{User defined relation editor}
%\vspace{-2em}

After the user is satisfied with the MRE matches, 
the user moves to define relations and code actions. 
The relation editor allows the user to define relations 
by specifying $\langle e_1,e_2,r\rangle$ tuples, 
where $e_1$ and $e_2$ denote source and destination entities, and $r$ denotes 
the label.
The editor shows the MRE tree and allows the user to select the sub 
expressions and select features 
of the matches of the sub expressions to define the three components of the relation. 
%The user first adds a relation, then assigns the values for the entities and relational entity (edge). 
%For each element in the tuple, the user specifies the sub-expression match and relevant label. 
%The label can be a match text, position, length, or numerical value, or a user-defined label.

%Recall the semantic relations we defined in the direction extraction task; namely
%relations $\langle e_1,e_2,r\rangle$, $\langle r,e_1,o_1\rangle$, and $\langle r,e_2,o_2\rangle$. 
Figure~\ref{fig:relationeditor} shows the declaration of the first relation 
$\langle s_9, s_{14},s_{3}\rangle$ where the 
labels of the nodes and the edges in the relational entities 
are defined as the text of the subexpression matches;
the editor allows the user to select the labels to be one of the match features. 
%was selected to be the entity node and relational entity edge labels. 
%Similarly, we define the other relations 
%$\langle s_{3},s_9,s_{10}\rangle$, and $\langle s_{3},s_{14},s_{11}\rangle$ 
%corresponding to $\langle r,e_1,o_1\rangle$, and $\langle r,e_2,o_2\rangle$, respectively.

%After specifying the semantic relations, we call the semantic relation constructor. 
A snapshot of the \framework GUI in 
Figure~\ref{fig:treegraph} shows shows in an interactive graph view
the entity-relation graph of the match of the user defined relation 
extracted from the match tree of the MRE. 
%The entity-relation graph is visualized when the user selects an MRE match tag from the tag list.
%The graph in Figure~\ref{fig:treegraph} shows the entity-relation graph of the match tree shown 
%in Figure~\ref{fig:treegraph}(b). 
%The interactive graph view enables the user to move the nodes.

%\subsection{Computational action editor}

### 2018-03-14 tex/merf/OldSubmission/introduction.tex
old submission files

{\em Computational Linguistics} (CL) is concerned with building 
accurate linguistic computational models. 
{\em Natural Language Processing} (NLP) is concerned with automating the 
understanding of natural language. 
CL and NLP tasks range from simple ones such as spell checking and typing 
error correction to more 
complex tasks including %text summarization, 
{\em named entity recognition} (NER), {\em cross-document analysis}, 
{\em machine translation}, 
and {\em relational entity extraction}~\cite{linckels2011natural,ferilli2011natural}.
Entities are elements of text that are of interest to an NLP task. 
Relational entities are elements that connect entities. 
{\em Annotations} relate chunks of text to {\em labels} denoting
semantic values such as entities or relational entities.
We refer to annotations and labels as {\em tags} and {\em tag types}, respectively, 
in the sequel.

Supervised and unsupervised empirical learning techniques tackle NLP and CL tasks. 
They employ machine learning without the need to manually encode the requisite 
knowledge~\cite{soudi2007arabic}. 
Supervised learning techniques require training corpora annotated 
with {\em correct} tags to learn a computational model. 
Supervised and unsupervised techniques require annotated reference corpora to 
evaluate the accuracy of the technique using metrics such as precision and 
recall~\cite{englishtreebank,arabictreebank,chinesetreebank}.

Researchers build %the required %seed 
training and reference corpora 
either manually, incrementally using learning techniques, or using knowledge-based 
annotation techniques that recognize and extract entities and relational 
entities from text. 
Knowledge-based techniques use linguistic and 
rhetorical domain specific knowledge encoded into sets of rules 
to extract entities and relational entities~\cite{soudi2007arabic}.
While existing annotation, entity, and relational entity 
extraction tools exist
\cite{chiticariu2010systemt% SystemT
,atzmueller2008rule%Textmarker
,urbain2012user%
,settles2011closing% dualist
,mmax2%MMAX
,brat% BRAT
},
most of them lack Arabic language support, and almost all of them
lack Arabic morphological analysis support
% that is key to Arabic NLP due to the rich morphological nature of Arabic
~\cite{habash2006arabic}. 
Fassieh~\cite{attia2009fassieh} is a {\em commercial} Arabic annotation tool
with morphological analysis support and text factorization. 
However, this tool lacks support for entity and relational entity extraction.

@@

In this paper, 
we present \framework, a {\em morphology-based entity and relational entity
extraction framework for Arabic text }. 
\framework provides a user-friendly interface where the user defines tag types 
and associates them with \framework formulae that are regular expressions over 
\framework Boolean formulae. 
Boolean formulae are terms, negations of terms, and disjunctions of terms. 
Terms are matches to Arabic morphological features including 
prefix, stem, suffix, part of speech (POS) tags, gloss tags, extended synonym tags, 
and semantic categories. 
We discuss the importance of the morphological features supported in 
\framework terms in Section~\ref{sec:morph}; in brief, morphological preprocessing
is key to Arabic NLP.% due to the richness of Arabic morphology.

We illustrate the target of \framework using a simple example. 
Consider the text in Figure~\ref{fig:intromotiv} which contains
directions to Dubai Mall~\footnote{text taken from the Dubai Mall website
~\url{http://www.thedubaimall.com/ar/}}.
The framed words in the text are entities referring to names of people 
($n_1,n_2,n_3$), 
names of places ($p_1,\dots,p_7$), 
relative positions ($r_1,\dots,r_4$), 
and numerical terms ($u_1,u_2$). 
We would like to extract those entities, and then extract
the relational entities forming the graph in Figure~\ref{fig:intromotiv} 
where vertices express entities, 
and edges represent the relational entities.
\framework allows a regular user to do that interactively
in four phases that include morphological analysis, 
entity extraction based on morphological features, 
relational entity extraction, and entity cross-reference.

\framework regular expressions support operators such as concatenation, 
zero or one, zero or more, one or more, up to $M\in\mathbb{N}^+$, and logical 
conjunction and disjunction. 
An editor allows the user to associate an action with each sub-expression.
The user specifies the action with C++ code and uses an API to access information related to the matches such as text, position, length, morphological features, and numerical value.

\framework takes an Arabic text, a set of user-defined \framework Boolean 
formulae, and a set of user-defined \framework regular expressions. 
\framework computes the morphological solutions of the words in the 
input text then computes matches to the Boolean formulae. 
\framework then generates a {\em non-deterministic finite state 
automata} (NDFSA) for each expression and simulates it with the 
sequence of Boolean formulae matches to compute the regular 
expression matches. 
\framework generates executable code for the actions associated with
the regular expressions, 
compiles, links, and executes the generated code 
as shared object libraries. 
Finally, \framework constructs the semantic relations and 
cross-reference between entities. 
\framework also provides visualization tools to present the matches, 
and estimate their accuracy with respect to reference tags.

\subsection{Existing annotation and entity extraction tools}

@@

Knowledge-based techniques such 
as~\cite{zaghouani2010adapting,traboulsi2009arabic} propose local grammars 
with morphological stemming to perform NER. 
%The work in~
\cite{ZaMaHaCicling2012Entity} presents a method for extracting entities, 
events, and relations amongst them from Arabic text using a hierarchy of manually built 
finite state machines driven by morphological features and graph 
transformation algorithms. 
Such techniques require advanced linguistic and programming expertise. 
QARAB is a question answering system that takes 
Arabic natural language queries and provides short answers~\cite{hammo2002qarab}.

Researchers also proposed systems for automatic IE based 
on user specifications. 
CPSL is a common pattern specification language for finite-state 
grammar~\cite{appelt1998common}. 
%\framework is an extension to CPSL with action execution and relation construction. 
%It defines three sections for declaration, rule definition, and macros. 
The work in \cite{chiticariu2010systemt} presents SystemT, a system based on 
an algebraic Approach to Declarative information extraction (IE). 
%It uses a declarative rule language and an optimizer. 
TEXTMARKER is a rule-based IE system designed to extract structured data from 
text \cite{atzmueller2008rule}. 
\cite{urbain2012user} presents a user-driven relational model 
requiring a user natural language query to extract entities and relational entities.

This work significantly extends the position 
paper~\cite{JaZaMatar} that allows for manual, and 
basic morphology annotation.
\framework enables a regular user to incrementally create 
complex annotations for Arabic text based on automatic 
extraction of morphological tags through a user friendly interactive interface. 
\framework has the following advantages.
\begin{itemize}
  \item \framework provides a novel and intuitive visual interface to build Boolean formulae over morphological features, 
    build regular expressions over the resulting Boolean formulae, and thereafter compute automatic tags.
  \item To our knowledge, this morphology-based framework  is the first for Arabic entity and relational entity extraction.
  \item \framework provides the user with the ability to rapidly create annotated Arabic text corpora with sophisticated morphology-based tags.
\end{itemize}

In \framework, we make the following contributions.
\begin{itemize}
  \item \framework enables the user to define relations in a simple manner
    and automatically detects relational entities matching the user defined relations. 
  \item \framework enables the user to associate subexpressions
    with code actions, and executes the code action 
    when a corresponding match is found. 
    It also provides an API to enable access to match 
    features such as text, position, length, numerical value, and morphological features.
  \item \framework enables the user to tag words based on a synonymic relation using the $Syn^k$ feature.
\end{itemize}

\input{directionMBFMRE}

In the rest of this paper,
Section~\ref{sec:motivation} presents a motivational example. 
Section~\ref{sec:morph} introduces Arabic morphological analysis 
and its important role in Arabic NLP. 
%We present preliminaries necessary to understand the work 
%in Section~\ref{sec:preliminaries}. 
Section~\ref{sec:framework} explains the \framework methodology. 
Section~\ref{sec:gui} presents the visual and user friendly interface of \framework. 
Section~\ref{sec:related} presents and discusses related work.
Section~\ref{sec:results} presents the evaluation results. 
Finally, we conclude and discuss future work in Section~\ref{sec:conclusion}.

>>>>

\IEEEPARstart{C}{\em omputational Linguistics} (CL) is concerned with building 
accurate linguistic computational models. 
{\em Natural Language Processing} (NLP) is concerned with automating the 
understanding of natural language. 
CL and NLP tasks range from simple ones such as spell checking and typing 
error correction to more 
complex tasks including %text summarization, 
{\em named entity recognition} (NER), {\em cross-document analysis}, 
machine translation, 
and {\em relational entity extraction}~\cite{linckels2011natural,ferilli2011natural}.
Entities are elements of text that are of interest to an NLP task. 
Relational entities are elements that connect entities. 
{\em Annotations} relate chunks of text to {\em labels} denoting
semantic values such as entities or relational entities.
We refer to annotations and labels as {\em tags} and {\em tag types}, respectively, 
in the sequel.

Supervised and unsupervised empirical learning techniques tackle NLP and CL tasks. 
They employ machine learning without the need to manually encode the requisite 
knowledge~\cite{soudi2007arabic}. 
Supervised learning techniques require training corpora that are annotated 
with {\em correct} tags to learn a computational model. 
Supervised and unsupervised techniques require reference corpora that are
annotated with correct tags to evaluate the 
accuracy of the technique in terms of metrics such as precision and 
recall~\cite{englishtreebank,arabictreebank,chinesetreebank}.

Researchers build the required %seed 
training and reference corpora 
either manually, incrementally using learning techniques, or using knowledge-based 
annotation techniques that recognize and extract entities and relational 
entities from text. 
Knowledge-based techniques use linguistic and 
rhetorical domain specific knowledge encoded into sets of rules 
to extract entities and relational entities~\cite{soudi2007arabic}.
While existing annotation, entity, and relational entity 
extraction tools exist
\cite{chiticariu2010systemt% SystemT
,atzmueller2008rule%Textmarker
,urbain2012user%
,settles2011closing% dualist
,mmax2%MMAX
,brat% BRAT
},
most of them lack Arabic language support, and almost all of them
lack Arabic morphological analysis support that is key to Arabic NLP
due to the rich morphological nature of Arabic~\cite{habash2006arabic}. 
Fassieh~\cite{attia2009fassieh} is a {\em commercial} Arabic annotation tool
with morphological analysis support and text factorization. 
Fassieh lacks support for entity and relational entity extraction.

@@

In this paper, 
we present a {\em morphology-based entity and relational entity
extraction framework for Arabic text } (\framework). 
\framework provides a user-friendly interface where the user defines tag types 
and associates them with \framework formulae that are regular expressions over 
\framework Boolean formulae. 
Boolean formulae are terms, negations of terms, and disjunctions of terms. 
Terms are matches to Arabic morphological features including 
prefix, stem, suffix, part of speech (POS) tags, gloss tags, extended synonym tags, 
and semantic categories. 
We discuss the importance of the morphological features supported in 
\framework terms in Section~\ref{s:s:morph}; in breif, morphological preprocessing
is key to Arabic NLP due to the richness of Arabic morphology.

We illusrate the target of \framework using a simple example. 
Given the text in Figure~\ref{fig:intromotiv} that contains
directions to Dubai Mall~\footnote{text taken from the Dubai Mall website
~\url{http://www.thedubaimall.com/ar/}}.
The framed words in the text are entities referring to names of people 
($n_1,n_2,n_3$), 
names of places ($p_1,\dots,p_7$), 
relative positions ($r_1,\dots,r_4$), 
and numerical terms ($u_1,u_2$). 
We would like to extract those entities, and then extract
the relational entities forming the graph in Figure~\ref{fig:intromotiv} 
where vertices express entities, 
and edges represent the relational entities.
\framework allows a regular user to do that interactively
in four phases that include morphological analysis, 
entity extraction based on morphological features, 
relational entity extraction, and entity cross-reference.

\framework regular expressions support operators such as concatenation, 
zero or one, zero or more, one or more, up to $M$, and logical 
conjunction and disjunction operations. 
\framework editor allows the user to associate an action with each 
\framework sub-expression. 
The user specifies the action with C++ code and uses the 
\framework API to access information related to 
the matches such as text, position, length, morphological features, 
and numerical values.

\framework takes an Arabic text, a set of user-defined \framework Boolean 
formulae, and a set of user-defined \framework regular expressions. 
\framework computes the morphological solutions of the words in the 
input text then computes matches to the Boolean formulae. 
\framework then generates a {\em non-deterministic finite state 
automata} (NDFSA) for each expression and simulates it with the 
sequence of Boolean formulae matches to compute the regular 
expression matches. 
\framework generates executable code for the actions associated with
the regular expressions, 
compiles, links, and executes the generated code 
as shared object libraries.

Finally, \framework constructs the semantic relations and 
cross-reference between entities. 
\framework also provides visualization tools to present the matches, 
and estimate their accuracy with respect to reference tags.

\subsection*{Existing annotation and entity extraction tools}

@@

Knowledge-based techniques such 
as~\cite{zaghouani2010adapting,traboulsi2009arabic} propose local grammars 
with morphological stemming to perform NER. 
The work in~\cite{ZaMaHaCicling2012Entity} presents a method for extracting entities, 
events, and relations amongst them from Arabic text using a hierarchy of manually built 
finite state machines driven by morphological features and graph 
transformation algorithms. 
Such techniques require advanced linguistic and programming expertise. 
QARAB is a question answering system that takes 
an Arabic natural language query and provides short answers for it~\cite{hammo2002qarab}.

Researchers also proposed systems for automatic IE based 
on user specifications. 
CPSL is a common pattern specification language for finite-state 
grammar~\cite{appelt1998common}. 
%It defines three sections for declaration, rule definition, and macros. 
The work in \cite{chiticariu2010systemt} presents SystemT, a system based on 
an algebraic Approach to Declarative information extraction (IE). 
%It uses a declarative rule language and an optimizer. 
TEXTMARKER is a rule-based IE system designed to extract structured data from 
text \cite{atzmueller2008rule}. 
The work in~\cite{urbain2012user} presents a user-driven relational model 
requiring a user natural language query to extract entities and relational entities.

\framework is an extension to CPSL with action execution and relation construction. 
\framework enables a regular user to incrementally create 
complex annotations for Arabic text based on automatic 
extraction of morphological tags through a user friendly interactive interface. 
\framework has the following advantages.
\begin{itemize}
  \item \framework provides a novel and intuitive visual interface to build Boolean formulae over morphological features, 
    build regular expressions over the resulting Boolean formulae, and thereafter compute automatic tags.
  \item Up to our knowledge, \framework  is the first morphology-based framework for Arabic entity and relational entity extraction.
  \item \framework provides the user with the ability to rapidly create annotated Arabic text corpora with sophisticated morphology-based tags.
\end{itemize}

In \framework, we make the following contributions.
\begin{itemize}
  \item \framework enables the user to define relations in a simple manner
    and automatically detects relational entities matching the user defined relations. 
  \item \framework enables the user to associate subexpressions
    with code actions, and executes the code action 
    when a corresponding match is found. 
    \framework provides an API to enable access to match 
    features such as text, position, length, numerical value, and morphological features.
  \item \framework enables the user to tag words based on a synonymic relation using the $Syn^k$ feature.
\end{itemize}

The work in this paper significantly extends the position 
paper~\cite{JaZaMatar} that allows for manual, and 
basic morphology annotation. 
The rest of this paper is organized as follows. 
Section~\ref{sec:motivation} presents a motivational example. 
Then, we present an overview of \framework in Section~\ref{sec:overview}. 
%We present preliminaries necessary to understand the work 
%in Section~\ref{sec:preliminaries}. 
Section~\ref{sec:framework} explains the methodology behind \framework. 
Section~\ref{sec:gui} presents the visual and user friendly aspects of \framework. 
In Section~\ref{sec:results}, we present our result in comparison to other tools. 
In Section~\ref{sec:related}, we present and discuss related work. 
Finally, we present our conclusion and future work in Section~\ref{sec:conclusion}.
